---
title: Exponential population growth  
author: |
 | Matthew Malishev^1^* 
 |  
 | _^1^ Department of Biology, Emory University, 1510 Clifton Road NE, Atlanta, GA, USA, 30322_
#bibliography:/Users/malishev/Documents/Melbourne Uni/Thesis_2016/library.bib
fontsize: 10
geometry: margin=1in
documentclass: article
linkcolor: blue
urlcolor: blue
citecolor: red
output:
  pdf_document:
    includes:
      in_header: # add .tex file with header content
    highlight: tango
    template: null
    toc: yes
    toc_depth: 4
    number_sections: false
    fig_width: 4
    fig_height: 5
    fig_caption: true
    df_print: tibble 
    citation_package: biblatex # natbib
    latex_engine: xelatex #pdflatex # lualatex
    keep_tex: true # keep .tex file in dir 
  html_document:
    highlight: tango
    code_folding: hide
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float: yes
  word_document:
    highlight: tango
    keep_md: yes
    pandoc_args: --smart
    #reference: mystyles.docx
    toc: yes
inludes:
  before_body: before_body.tex
subtitle: 
tags:
- nothing
- nothingness
params: 
  dir: "/Users/malishev/Documents/Emory/lab/interns/sally/499R/"
  date: !r Sys.Date()
  session: !r sessionInfo()  
  version: !r getRversion()
  email: "matthew.malishev@gmail.com"
  doi: https://github.com/darwinanddavis
classoption: portrait
# ^['https://github.com/darwinanddavis'] # footnote
vignette: >
  %\VignetteIndexEntry{Useful R code}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

<script type="text/x-mathjax-config">
  MathJax.Hub.Config({ TeX: { equationNumbers: {autoNumber: "all"} } });
</script>

```{r echo = FALSE}
# library(rmarkdown)
# setwd("")
# f <- list.files()[1]
# render(f, output_format='pdf_document')
```

```{r, set-options, echo = FALSE, cache = F}
options(width=100)
knitr::opts_chunk$set(
 eval = F, # run all code
 # echo = FALSE, # show code chunks in output 
 comment = "",
 tidy.opts=list(width.cutoff=100), # set width of code chunks in output
 tidy=TRUE, # make output as tidy
 message = FALSE,  # mask all messages
 warning = FALSE, # mask all warnings 
 size="small" # set code chunk size
)

# https://github.com/ucb-stat133/stat133-fall-2016/blob/master/hws/hw02-tables-ggplot.Rmd
knitr::opts_knit$set(root.dir=paste0(params$dir,"/")) # set working dir

setwd(paste0(params$dir,"/")) # for running just in R not knitr
```

\newpage   

Date: `r params$date`  
R version: `r params$version`  
*Corresponding author: `r params$email`  
This document can be found at `r params$doi`  

\  

R session info 

```{r eval = T}
params$session
```      

\newpage  

```{r results="hide",echo=F}
require(RCurl)
script <- getURL("https://raw.githubusercontent.com/darwinanddavis/499R/master/exp_pop_growth.Rmd", ssl.verifypeer = FALSE)
eval(parse(text = script))
```


# Overview  

Examples of exponential population growth in R.   

## Install dependencies
```{r, load packages, include=T, cache=F, message=F, warning=F, results='hide'}
packages <- c("dplyr","deSolve","pdftools")  
if (require(packages)) {
    install.packages(packages,dependencies = T)
    require(packages)
}
lapply(packages,library,character.only=T)
```
# Section 1

## Exponential growth equation    

$$
N_t = N_0 \cdot e^{rt}
$$
Nt = the number of individuals in the population after t units of time  
No = the initial population size (t = 0)  
r = the exponential growth rate  
t = time unit (usually in years)  
e = the base of the natural logarithms (2.72)  

Exponential rate of growth is commonly named the parameter lambda $\lambda$    

$$
\lambda = e^{r}
$$

$e^r$ = lambda. Exponential growth rate parameter.  
\  
The natural log (ln) of e = 1

$$
ln(e) = 1
$$ 

because $e^1 = e$.      

\  

The natural log of 1 = 0

$$
ln(1) = 0
$$    

because $e^0 = 1$.  
  

### Parameters  
```{r}
# parameters 
N_t <- 0 # expected pop size
N_0 <- 500 # initial pop size
e <- exp
r <- 0.012 # exponetial rate of growth
lambda <- e(1^r)
t <- 10 # time (in years)
```



```{r}
# putting the above all together in R  
N_t <- N_0 * e(r*t)
N_t
```

### Example  

A moose population has a growth rate of 0.02. In 2000, the population
was 500. What will the population be in 2020?

```{r}
# input your R code here  

```


## Instantaneous rate of growth  
Equation showing the rate of population increase  



$$
\frac{dN}{dt} = rN
$$
dN = change in number  
dt = change in time  
r = the per head maximum potential growth rate  
N = number of individuals in a population  

```{r}
# in R
N <- 1000
dNdt <- r*N
dNdt
```
### Example  

A population of 100 individuals. Each individual can on average contribute 1/4 of an individual (new individual) to the population in a given unit of time. Find the rate of population increase.   

```{r}
# your r code  
```

## Simulating population growth  

Set your parameters for the population  
```{r}
N_0 <- 20; # initial population size  
```

Over time  

```{r}
N_1 <-  N_0 * r; # population size at t = 1  
```

What does this look like at each time point?  

```{r}
N_2 <- # ??
N_3 <- # ??  
# etc
```

Population size     

```{r}
popsize = c(N_0, N_1, N_2, N_3, N_4, N_5)
popsize  
popsize[2]  
```
# Section 2  

## Simulation loop   

```{r}
r <- 2
N_0 <-  20
t <- 10
popsize <- list() # create empty list to populate in loop  
tt <- 2:t # now make t a vector to make a time period

for(i in tt){ # start  loop
  N_0[i] <- N_0[i-1]*r # middle of loop
  popsize <- N_0 # put the pop size into the list we created  
} # end loop
popsize
```

Plot  
```{r}
# plot it
plot(popsize)

# extend plot params
tta <- c(1,tt)# add the first year onto the time vector
ttm <- max(tt) # get the maximum value of the time vector  
xlim <- c(1,ttm) # put this in a vector
ylim <- c(0,max(N_0)) # do the same for growth
par(las=1,bty="n") # set plotting params  

plot(tta,popsize, # vectors to plot
     xlim=xlim, # set the x limits
     ylim=ylim, # y limits  
     xlab="Time",
     ylab="Population growth",
     pch=20, # set point type
     col="pink"
     )

?plot # help page for plot function   

```

## Event-based conditions   

Changing conditions halfway through a simulation. Using `if` and `else` statements.  
```{r}
popsize <- list() # reset the list  

for (i in tt){ 
  if((i%%5) == 0){ #modify growth rate depending on the year
    r <- 4
    }else{ # or just an ok year
      r <- 2
      }
  N_0[i] <- N_0[i-1]*r # the population growth equation
  popsize <- N_0
} 

popsize

par(las=1,bty="n") # set plotting params   
plot(popsize,pch=20,col="pink")
```
Another `for` loop example  

```{r}
popsize <- list() # reset the list  

for (i in tt){ 
  if (i <= 6){ # for the first five years, use r = 2
    r <- 2
    }else if(i >= 10){
      # and also at 10 years 
      r <- 2
      }else{
    # but between 5th and 9th year,
    # we grow slower
        r <- 0.5
        } # end of if elseif else statement
  N_0[i] <- N_0[i-1]*r # the population growth equation
  popsize <- N_0
} 

popsize

par(las=1,bty="n") # set plotting params   
plot(popsize,pch=20,col="pink")
```

How do we figure out at what time point a population has reached a certain growth stage? Using `while` loops.    
```{r}

r <- 2 # ?? # set r
N_0 <-  20 # initial size
popsize <- list() # reset the list  
popsize_d <- 5000 # desired pop size

t <- 0 # set time var 
while (N_0[t + 1] <= popsize_d){ # while we're under the desired pop size  
   t = t + 1 ; # advance one time step  
  N_0[t + 1] <- N_0[t]*r; # population growth eq
  time_to_popsize_d <- NULL # time to reach popsize_d? 
  popsize <- N_0
} # close loop

```

How can we use `R` to get the time to reach `popsize_d`?    
```{r}

# get the length of the populaiton size vector and replace the "NULL" in the above loop with this function.  
time_to_popsize_d <- NULL # ??

popsize[time_to_popsize_d] # this is the size of the pop at the desired time step  

```

## Making reusable `R` code: functions.    

Putting all the above into reusable code that replaces the population growth parameters with input variables. Using `function`.  

`Function`s pass user-defined input variables into an `R` procedure so that you can package code into a reusable and general program, the function.     

Take your simulation you want to put into the function.  
```{r}
r <- 2 # set r
popsize <-  20 # initial size
popsize_d <- 5000 # desired size
time_to_popsize_d <- NULL # ?? # use your answer from above 

t <- 0 # set time var 
while (popsize[t + 1] <= popsize_d){ # while we're under the desired pop size  
   t = t + 1 ; # advance one time step  
  popsize[t + 1] <- popsize[t]*r; # population growth eq
  time_to_popsize_d <- NULL # time to reach popsize_d?? 
} # close loop

# writing the function. we'll call it 'discrete_growth'. 
discrete_growth <- function(r){ # make popsize an input variable   
  d <- popsize*r
  return(d) # we're just returning the popsize multipled by growth rate
} 

# now we replace the 'r' parameter with a value for 'r'
discrete_growth(5)

```

What would the function look like if we wanted to figure out the following parameters?
```{r}
r <- NULL #?? pop growth 
N_0 <- NULL #?? initial pop size  
t <- NULL
popsize <- list() # create empty list to populate in loop  

# write the below function  
discrete_growth <- function(??){ # <--- there will be three input variables  
  popsize <- ??
  tt <- 2:?? # make t a vector to make a time period
  for (i in tt){
    N_0[i] <- N_0[i-1]*r # middle of loop
    popsize <- N_0 # put the pop size into the list we created  
    } # end loop
  d <- popsize 
  plot(d) # plot popsize with the appropriate time scale  
  return(d) # get d
} 

# now just replace the input variable with the values you want to simulate  
discrete_growth() 

```

Save the function in the `R` memory. Call it 'popsize_func'.  
```{r}
popsize_func <- ??

```

Now save a separate function with different input parameters, then run both functons. This should give you an idea of how `R` saves things to memory.   
```{r}
# use these values for popsize_func_2
0.1
500
10000

popsize_func_2 <- ?? # 
popsize_func # is the output for popsize_func look different to popsize_func_2? 
```

